Draft -- Discriminating Seagrasses From Green Macroalgae in European Intertidal areas using high resolution multispectral drone imagery -- Draft

### 2024-11-14 docs/_tex/index.tex
add laurent's comments

\begin{abstract}
Coastal areas support seagrass meadows, which offer crucial ecosystem
services including erosion control and carbon sequestration. However,
these areas are increasingly impacted by human activities, leading to
habitat fragmentation and seagrass decline. In situ surveys,
traditionally performed to monitor these ecosystems face limitations on
temporal and spatial coverage, particularly in intertidal zones,
prompting the addition of satellite data within monitoring programs.
Yet, satellite remote sensing can be limited by too coarse spatial
and/or spectral resolution, making it difficult to discriminate seagrass
from other macrophytes in highly heterogenous meadows. Drone (unmanned
aerial vehicles -- UAV) images at a very high spatial resolution offer a
promising solution to address challenges related to spatial
heterogeneity and intrapixel mixture. This study focuses on using drone
acquisitions with a ten spectral band sensor similar to that onboard
Sentinel-2, for mapping intertidal macrophytes at low tide (i.e.~during
a period of emersion) and effectively discriminating between seagrass
and green macroalgae. Nine drone flights were conducted at two different
altitudes (12 m and 120 m) across heterogeneous intertidal European
habitats in France and Portugal, providing multispectral reflectance
observation at very high spatial resolution (8 mm and 80 mm
respectively). Taking advantage of their extremly high spatial
resolution; the low altitude flights were used to train a Neural Network
classifier to discrimintate five taxonomic classes of intertidal
vegetation: Magnoliopsida (Seagrass), Chlorophyceae (Green macroalgae),
Phaeophyceae (Brown algae), Rhodophyceae (Red macroalgae) and benthic
Bacillariophyceae (Diatoms), and validated using concomitant field
measurements. Classification of drone imagery resulted in an overall
accuracy of 94\% across all sites and images, covering a total area of
467 000 m². The model exhibited an accuracy of 96.4\% in identifying
seagrass. In particular , seagrass and green algae can be discriminated.
The very high spatial resolution of the drone data made it possible to
assess the influence of spatial resolution on the classification
outputs, showing a limited loss in seagrass detection up to about 10m.
Altogether, out findings suggest that the Multi-Spectral Instrument
(MSI) onboard Sentinel-2 offers a relevant trade-off between its spatial
and spectral resolution, thus offering promising perspectives for
satellite remote sensing of intertidal biodiversity over lager scales.
\end{abstract}

@@

Drones (Unmanned Aerial Vehicles -- UAVs) can potentially fill the data
gaps left by satellite RS and \emph{in situ} measurements, due to their
ability to provide spatially-explicit observations at very high spatial
resolutions (pixel size from mm to cm) while capturing data at
multi-spectral resolution \citep{fairley2022drone, oh2017use}. The
versatility of drones allows for their application across a diverse
thematic range , from coastal zone management
\citep{adade2021, casella2020, angnuureng2022} to mapping species
distribution
\citep{joyce2023, tallam2023, Roca2022, Roman2021, Brunier2022Topographic, sousa2019blue}.
However, when applied to coastal habitat mapping, previous case studies
were mostly limited to a low number of \textbf{drone} flights over a
single study site, restricting the generalizability of their application
over wider geographical scales
\citep{Roman2021, collin2019improving, rossiter2020uav, Brunier2022Topographic}.
\textbf{These studies have demonstrated the capability of drones to map
intertidal habitats, including seagrasses; however, broader
generalization of these findings is still lacking. The current paper
uniquely expands the spatial and methodological scope of drone-based
remote sensing for intertidal habitat mapping across a broad
biogeographical range. It demonstrates the feasibility of accurately
classifying diverse macrophyte types across various study sites, with a
particular focus on distinguishing Magnoliopsida (seagrasses) and
Chlorophyceae (green algae). Unlike previous studies, our approach
integrates multiple spatial scales by simulating satellite resolutions
and quantifying the impact of spatial resolution on classification
accuracy.} Nine drone flights were performed over soft-bottom intertidal
areas along the Atlantic coastlines of two European countries (France
and Portugal), covering a wide range of habitats, from monospecific
seagrass meadows to meadows mixed with green, or red macroalgae. A deep
learning algorithm was trained and validated for macrophyte
discrimination, emphasizing applicability across diverse sites without a
loss of prediction accuracy. The classification maps obtained at a very
high spatial resolution with the drone were spatially degraded to
satellite resolutions, making it possible to assess the effect of
spatial resolution on classification accuracy, and provide insights for
coastal habitat mapping using satellite remote sensing. \textbf{This
study is, therefore, among the first to quantify the effects of spatial
resolution on the accuracy of drone-based macrophyte classification
across a wide geographical scale, providing a framework for to better
understand satellite-based classification challenges.}

@@

\caption{\label{fig-map}Location of drone flights in France and
Portugal. A: Gulf of Morbihan (Two sites), B: Bourngeuf Bay (Two sites),
C: Ria de Aveiro Coastal Lagoon (Three sites). Green areas represents
the intertidal zone.}

@@

At each location, a DJI Matrice 200 quadcopter drone equipped with a
Micasense RedEdge Dual MX multispectral camera was flown to take 1.2
million pixel reflectance photographs with ten spectral bands ranging
from the blue to the near infrared (NIR): 444, 475, 531, 560, 650, 668,
705, 717, 740 and 840 nm. To ensure consistent lighting conditions
across flight paths, the drone's trajectory was aligned to maintain a
solar azimuth angle of 90 degrees. An overlap of 70\% and 80\% (side and
front respectively) between each image was set for each flight. A
downwelling light sensor (DLS2) was used to acquire irradiance data
concomitantly with the camera measurements. Raw data were calibrated in
reflectance using a calibration panel reflective at \textasciitilde50\%
provided by the manufacturer. Across all sites, flights were made at two
different altitudes : 12 m or/and 120 m, with a spatial resolution of 8
mm and 80 mm, respectively (Table~\ref{tbl-flights}). \textbf{Low
altitude flights were used as training dataset for a neural network
model while high altitude flight were used for validation.}

@@

\caption{\label{tbl-flights}List of drone flight, summarising the date,
the altitude, and the purpose of each flight. 12 m and 120 m flights
have a spatial resolution of 8 and 80 mm respectively.}

@@

\includegraphics[width=6.63in,height=\textheight,keepaspectratio]{Figures/High_res/Table1/table_flights.png}

@@

Before each flight, targets used as ground control points were
distributed over the study site and georeferenced with a Trimble © Geo
XH 6000 differential GPS (dGPS). Ground control points were used to
correct georeferencing imprecision of orthomosaics with an horizontal
and vertical accuracy of 10cm. A dGPS was also used to georeference
quadrats of 0.25 m² ,which assessed the presence or absence of five key
taxonomic classes of intertidal vegetation : Bacillariophyceae
(unicellular benthic diatoms forming biofilms at the sediment surface
during low tide, \textbf{ranging in size from small patches (m²) to
entire mudflats (km²)}; henceforth: Diatoms), Phaeophyceae (henceforth:
brown macroalgae), Magnoliopsida (henceforth: Seagrasses), Chlorophyceae
(henceforth: Green macroalgae) and Rhodophyceae (henceforth: red
macroalgae) (Figure~\ref{fig-vegetation}). Only homogeneous vegetation
patches extending over several meters were selected as ground control
points. Pictures of each quadrat were uploaded online to the open-portal
Global Biodiversity Information Facility (GBIF) platform
\citep{BedeGbif}. Each photograph was also processed to estimate the
percent cover of each type of vegetation using an image processing
software \citep[ImageJ,][]{schneider2012nih}. Hyperspectral reflectance
signatures of each vegetation class were recorded using an ASD FieldSpec
HandHeld 2 spectroradiometer, which acquires reflectance between 325 and
1075 nm, with 1 nm of spectral resolution. Hyperspectral signatures
served dual purposes: they validate the radiometric calibration of drone
data and contribute to misclassification reduction in photo
interpretations.

@@

\caption{\label{fig-vegetation}The five taxonomic classes of vegetation
used to train the Neural Network model and an example of their raw
spectral signatures at the spectral resolution of the Micasense RedEdge
Dual MX. A : Magnoliopsida (\emph{Zostera noltei}) ; B : Phaeophyceae
(\emph{Fucus sp.}) ; C : Rhodophyceae (\emph{Gracilaria
vermiculophylla}) ; D : Chlorophyceae (\emph{Ulva sp.}) ; E :
Bacillariophyceae (Diatoms - MPB). \textbf{Classes were identified
following the World register of Marine species.}}

@@

A structure-from-motion photogrammetry software \citep[Agisoft
Metashape,][]{agisoft} was used to process images to obtain
multispectral orthomosaics of each flight. The process for
orthomosaicking was identical for every flight. First, key tying points
were detected inside of each image and between overlapping images in
order to obtain a sparse point cloud. This cloud was cleaned using a
reprojection accuracy metric in order to remove noisy points. A dense
point cloud was then produced using a structure from motion algorithm. A
surface interpolation of this dense point cloud was made to obtain a
digital surface model (DSM), used to reconstruct the multispectral
ortho-image \citep{nebel2020review}. Low altitude drone flights produced
ortho-images with a very high spatial resolution (8 mm per pixel),
making it efficient to visually distinguish between the various types of
vegetation. High altitude flights allowed to cover larger areas and
produced images with a pixel size of 80 mm (Table~\ref{tbl-flights}).

@@

The spectral similarities of the reflectance signatures at the spectral
resolution of the micasense senor between intertidal green macrophytes
(Magnoliopsida and Chlorophyceae) make their discrimination challenging
using simple classification algorithms (Figure~\ref{fig-vegetation} F).
To overcome this challenge, a deep learning classification method was
trained, validated, and applied to each drone flight
(Figure~\ref{fig-workflow}).

@@

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./Figures/Low_res/Figure3_workflow.png}

@@

\caption{\label{fig-workflow}Schematic representation of the workflow.
Parallelograms represent input or output data, and rectangles represent
Python processing algorithms. The overall workflow of this study is
divided into two distinct parts based on the spatial resolution of the
drone flights: high-resolution flights (pixel size: 8 mm) were utilized
for training and prediction of the Neural Network model, whereas
lower-resolution flights (pixel size: 80 mm) were solely employed for
prediction purposes. Validation has been performed on both high and low
resolution flights.}

@@

A dataset containing photo-interpreted drone reflectance pixels was
built to train a Neural Network model. The training pixels were
categorized into seven different classes, representing the various
habitats encountered at the different study sites: Sediment, Water,
green macroalgae, seagrasses, diatoms, brown macroalgae and red
macroalgae. Only data from the low-altitude flights
(Table~\ref{tbl-flights}) were used for training because their 8 mm
spatial resolution allowed to avoid spectral sub-pixel mixing and to
accurately identify vegetation classes. \textbf{In the field, seagrasses
displaying two distinct colors, each with a unique spectral signature,
were observed. Careful attention was given to incorporating training
pixels from both color types into the training dataset for the seagrass
class. This approach was consistently applied to all classes within the
model.} More than 418,000 pixels at 8 mm resolution from the 3 training
flights were used to train the model (Table~\ref{tbl-validationPX}).
Twenty one variables were used by the model as predictors: the ten raw
spectral bands of the Micasense RedEdge Dual MX multispectral camera
(ranging from 444 nm to 840 nm), the same ten spectral bands
standardized using a min/max transformation (Equation~\ref{eq-std} ;
\citep{Cao2017}) and the Normalized difference vegetation index (NDVI,
Equation~\ref{eq-ndvi}). Standardisation of spectral bands is commonly
used to eliminate the scaling differences between spectra and to limit
the effect of biomass on the spectra shape
\citep{Douay2022, Davies2023}.

@@

A neural network classification model was built using the fastai
workflow \citep{howard2018fastai}. This model was composed of 2 hidden
layers and have a total of 26 054 trainable parameters. Parameters have
been fine tuned using 12 epoch to minimize the error rate. This model
has been called DISCOV, standing for Drone Intertidal Substrat
Classification Of Vegetation.

@@

The workflow of this study revolves around two distinct flight heights
(12 and 120 m, Figure~\ref{fig-workflow}) where ensuring consistency
between reflectances at both heights is crucial. This comparison was
conducted at sites where low and high altitude flights overlapped. The
low altitude flights were resampled to the same spatial resolution and
grid as the high flights using a median resampling method. Reflectance
values were then extracted, and a scatterplot was generated and the Root
Mean Square Error (RMSE) was computed to compare the difference bewteen
the raw and standardised reflectance.

@@

Variable Importance Plots (VIP) serve as a method to identify which
predictors are important for predicting a specific class. Out of the 21
predictors utilized in this study, Variable Importance was computed only
for the raw and standardized values of the 10 spectral bands captured by
the MicaSense camera. This is achieved by repeatedly predicting the same
dataset while randomly shuffling one predictor at a time. The benchmark
score obtained after each iteration is then compared to the benchmark
score obtained without shuffling any variables. The greater the
difference between these two benchmark values, the more important the
variable is for the model \citep{WEI2015399}.

@@

We used a Generalized Linear Model (GLM) with a Beta distribution to
examine the relationship between pixel resolution, vegetation class, and
their interaction on the loss of vegetation. The loss of vegetation was
modelled as function of the interaction between pixel resolution and
vegetation class (Diatoms, brown macroalgae, seagrass, green macroalgae
and red macroalage). Sample vs fitted residuals and quartile-quartile
graphics were assessed visually, to ensure assumption of the models used
were met.

@@

The key aspect of the workflow adopted in the present study is the
mapping at two different altitudes (12 and 120 m), resulting in two
distinct resolutions for the same area (8 and 80 mm; respectively). The
high-resolution flight was used to estimate the sub-pixel composition
for each pixel of the lower-resolution flight. Consequently, within each
pixel of the high-altitude flights, the contribution of each vegetation
class (\% cover) was obtained, and a kernel density plot was generated.
This plot provided a visual representation of the model's behavior in
mixed vegetation scenarios. It helps to understand the minimum
vegetation cover of a given class within a pixel necessary for the model
to confidently predict that class.

@@

In this study, drone flights were conducted at two different altitudes
(12 and 120 m) to construct the neural network model. At the sites where
the flights at both altitudes overlapped, reflectance was compared.
Overall there was a good agreement between the two altitudes (RMSE :
0.027 ; Figure~\ref{fig-CompareRef}).

@@

There was a slight underestimation for raw reflectance values in the
high-altitude flight, particularly for higher reflectance values
(Figure~\ref{fig-CompareRef} A). Since both flights were conducted over
vegetated areas, the highest reflectance values correspond to the
infrared part of the spectrum. This difference was not present when the
reflectance has been standardized (Equation~\ref{eq-std} ;
Figure~\ref{fig-CompareRef} B).

@@

Each drone flight was used to produce a prediction map, as well as a
probability map that indicates the model derived probability of the
selected class for every pixel. The low-altitude flight conducted in
Gafanha, Portugal, represented the site with the highest complexity
(Figure~\ref{fig-GafLow}). Among the five vegetation classes on which
the model was trained, four were present on this site, with green and
red macroalgae mixed with a seagrass meadow. There was also diatoms
forming biofilms on bare sediment surface. Although the seagrass was
solely composed of a single specie, \emph{Zostera noltei}, various
colors of this specie could be observed from dark green (corresponding
to healthy leaves) to brown (when leaves are senescent or have an
altered pigment composition). Regardless of the variation of color, the
class Magnoliopsida (seagrass) was accurately predicted by the model (F1
score of 0.96 at that site).

@@

\caption{\label{fig-GafLow}RGB orthomosaic (Left) and Prediction (Right)
of the low altitude flight of Gafanha, Portugal. The total extent of
this flight was 3000 m² with a resolution of 8 mm per pixel. Background
colors indicate intertidal area (Light Green) and land area (Light
Grey). The zoom covers an area equivalent to a 10-meter Sentinel-2 pixel
size.}

@@

The high-altitude flight over Gafanha covered a total area of
\textasciitilde1 km² (Figure~\ref{fig-GafHigh}). A channel contouring a
small island was masked in the prediction map. Most of vegetation area
was classified as seagrass by the model, including patches with
discolored leaves. Only a few pixels were classified as green macroalgae
(F1 score of 0.55). Patches of red macroalgae were correctly classified
(F1 score of 0.85). In the northern part of the site and near the land
eadges, patches of the schorre angiosperm \emph{Sporobolus maritimus}
(syn. \emph{Spartina maritima)} were misclassified, either as seagrass
or as brown algae (F1 score of 0.77 and 0.71, respectively).

@@

\caption{\label{fig-GafHigh}RGB orthomosaic (Left) and Prediction
(Right) of the high altitude flight of Gafanha, Portugal. The total
extent of this flight was about 1 km² with a resolution of 80 mm per
pixel. Background colors indicate intertidal area (Light Green), land
area (Light Grey) and water (Light Blue). The yellow outline shows the
extent of the low altitude flight of Gafanha presented in
Figure~\ref{fig-GafLow}. The zoom covers an area equivalent to a
10-meter Sentinel-2 pixel size.}

@@

Among the high altitude flights, the one acquired over the inner part of
Ria de Aveiro coastal lagoon covered the largest area with approximately
1.5 km² (Figure~\ref{fig-Boat}). Teh vegetation present at the site was
dominated by seagrass and red macroalgae. The classification provided
consistent results, with a patchy seagrass meadow mixed with red
macroalgae on the eastern part of the site. As shown in the zoom
(Figure~\ref{fig-Boat}), the edges of the meadow were mixed with green
macroalgae (\emph{Ulva sp.}), which the model agreed with (F1 score of
0.89 for green algae, 0.97 for seagrass and 0.98 for red algae).

@@

\caption{\label{fig-Boat}RGB orthomosaic (Top) and Prediction (Bottom)
of the flight made in the inner part of Ria de Aveiro Lagoon, Portugal.
The total extent of this flight was about 1.5 km² with a resolution of
80 mm per pixel. Background colors indicate intertidal area (Light
Green), land area (Light Grey) and water (Light Blue). The zoom inserts
cover an area equivalent to the size of a 10-meter Sentinel-2 pixel.}

>>>>

\begin{abstract}
Coastal areas support seagrass meadows, which offer crucial ecosystem
services including erosion control and carbon sequestration. However,
these areas are increasingly impacted by human activities, leading to
habitat fragmentation and seagrass decline. In situ surveys,
traditionally performed to monitor these ecosystems face limitations on
temporal and spatial coverage, particularly in intertidal zones,
prompting the addition of satellite data within monitoring programs.
Yet, satellite remote sensing can be limited by too coarse spatial
and/or spectral resolution, making it difficult to discriminate seagrass
from other macrophytes in highly heterogenous meadows. Drone (unmanned
aerial vehicles -- UAV) images at a very high spatial resolution offer a
promising solution to address challenges related to spatial
heterogeneity and intrapixel mixture. This study focuses on using drone
acquisitions with a ten spectral band sensor similar to that onboard
Sentinel-2, for mapping intertidal macrophytes at low tide (i.e.~during
a period of emersion) and effectively discriminating between seagrass
and green macroalgae. Nine drone flights were conducted at two different
altitudes (12 m and 120 m) across heterogeneous intertidal European
habitats in France and Portugal, providing multispectral reflectance
observation at very high spatial resolution (8 mm and 80 mm
respectively). Taking advantage of their extremly high spatial
resolution; the low altitude flights were used to train a Neural Network
classifier to discrimintate five taxonomic classes of intertidal
vegetation: Magnoliopsida (Seagrass), Chlorophyceae (Green macroalgae),
Phaeophyceae (Brown algae), Rhodophyceae (Red macroalgae) and benthic
Bacillariophyceae (Benthic diatoms), and validated using concomitant
field measurements. Classification of drone imagery resulted in an
overall accuracy of 94\% across all sites and images, covering a total
area of 467 000 m². The model exhibited an accuracy of 96.4\% in
identifying seagrass. In particular , seagrass and green algae can be
discriminated. The very high spatial resolution of the drone data made
it possible to assess the influence of spatial resolution on the
classification outputs, showing a limited loss in seagrass detection up
to about 10m. Altogether, out findings suggest that the Multi-Spectral
Instrument (MSI) onboard Sentinel-2 offers a relevant trade-off between
its spatial and spectral resolution, thus offering promising
perspectives for satellite remote sensing of intertidal biodiversity
over lager scales.
\end{abstract}

@@

Drones (Unmanned Aerial Vehicles -- UAVs) can potentially fill the data
gaps left by satellite RS and \emph{in situ} measurements, due to their
ability to provide spatially-explicit observations at very high spatial
resolutions (pixel size from mm to cm) while capturing data at
multi-spectral resolution \citep{fairley2022drone, oh2017use}. The
versatility of drones allows for their application across a diverse
thematic range , from coastal zone management
\citep{adade2021, casella2020, angnuureng2022} to mapping species
distribution
\citep{joyce2023, tallam2023, Roca2022, Roman2021, Brunier2022Topographic, sousa2019blue}.
However, when applied to coastal habitat mapping, previous case studies
were mostly limited to a low number of \textbf{drone} flights over a
single study site, restricting the generalizability of their application
over wider geographical scales
\citep{Roman2021, collin2019improving, rossiter2020uav, Brunier2022Topographic}.
\textbf{These studies have demonstrated the capability of drones to map
intertidal habitats, including seagrasses; however a broader
generalization of these findings is still lacking. The current paper
uniquely expands the spatial and methodological scope of drone-based
remote sensing for intertidal habitat mapping across a broad
biogeographical range. It demonstrates the feasibility of accurately
classifying diverse macrophyte types across various study sites, with a
particular focus on distinguishing Magnoliopsida (seagrasses) and
Chlorophyceae (green algae). Unlike previous studies, our approach
integrates multiple spatial scales by simulating satellite resolutions
and quantifying the impact of spatial resolution on classification
accuracy.} Nine drone flights were performed over soft-bottom intertidal
areas along the Atlantic coastlines of two European countries (France
and Portugal), covering a wide range of habitats, from monospecific
seagrass meadows to meadows mixed with green, or red macroalgae. A deep
learning algorithm was trained and validated for macrophyte
discrimination, emphasizing applicability across diverse sites without
losing prediction accuracy. The classification maps obtained at a very
high spatial resolution with the drone were spatially degraded to
satellite resolutions, making it possible to assess the effect of
spatial resolution on classification accuracy, and provide insights for
coastal habitat mapping using satellite remote sensing. \textbf{This
study is, therefore, among the first to quantify the effects of spatial
resolution on the accuracy of drone-based macrophyte classification
across a wide geographical scale, providing a framework to better
understand satellite-based classification challenges.}

@@

\caption{\label{fig-map}Location of drone flights in France and
Portugal. A: Gulf of Morbihan (Two sites), B: Bourgneuf Bay (Two sites),
C: Ria de Aveiro Coastal Lagoon (Three sites). Brown areas represents
the intertidal zone.}

@@

At each location, a DJI Matrice 200 quadcopter drone equipped with a
Micasense RedEdge Dual MX multispectral camera was flown to take 1.2
million pixel reflectance photographs with ten spectral bands ranging
from the blue to the near-infrared (NIR): 444, 475, 531, 560, 650, 668,
705, 717, 740 and 840 nm. To ensure consistent lighting conditions
across flight paths, the drone's trajectory was aligned to maintain a
solar azimuth angle of 90 degrees. An overlap of 70\% and 80\% (side and
front respectively) between each image was set for each flight. A
downwelling light sensor (DLS2) was used to acquire irradiance data
concomitantly with the camera measurements. Raw data were calibrated in
reflectance using a calibration panel reflective at \textasciitilde50\%
provided by the manufacturer. Across all sites, flights were made at two
different altitudes : 12 m or/and 120 m, with a spatial resolution of 8
mm and 80 mm, respectively (Table~\ref{tbl-flights}).
\textbf{Low-altitude flights, with a spatial resolution of 8 mm, were
used to build the training dataset for the neural network, as this high
resolution allowed for precise photo-interpretation of vegetation
classes. In contrast, high-altitude flights were used for validation
purposes.}

@@

\caption{\label{tbl-flights}List of drone flights, summarising the date,
the altitude, and the purpose of each flight. 12 m and 120 m flights
have a spatial resolution of 8 and 80 mm respectively.}

@@

\includegraphics[width=8.84in,height=\textheight,keepaspectratio]{Figures/High_res/Table1/table_flights.png}

@@

Before each flight, targets used as ground control points were
distributed over the study site and georeferenced with a Trimble © Geo
XH 6000 differential GPS (dGPS). Ground control points were used to
correct georeferencing imprecision of orthomosaics with an horizontal
and vertical accuracy of 10cm. A dGPS was also used to georeference
quadrats of 0.25 m², which assessed the presence or absence of five key
taxonomic classes of intertidal vegetation: Bacillariophyceae
\textbf{(benthic diatoms forming biofilms at the sediment surface during
low tide with biofilm's size ranging from small patches (m²) to entire
mudflats (km²); henceforth: Benthic diatoms), Phaeophyceae (brown
macroalgae generally attached to rocks or other substrates able to form
dense beds in the intertidal zone; henceforth: Brown macroalgae),
Magnoliopsida (seagrasses, rooted flowering marine plants able to form
extensive meadows on soft sediments; henceforth: Seagrasses),
Chlorophyceae (green macroalgae, typically found attached to rocks or
washed ashore; henceforth: Green macroalgae), and Rhodophyceae (red
macroalgae, attached to hard substrates but can also be found on
soft-bottom substrate; henceforth: Red macroalgae)}. Only homogeneous
vegetation patches extending over several meters were selected as ground
control points. Pictures of each quadrat were uploaded online to the
open-portal Global Biodiversity Information Facility (GBIF) platform
\citep{BedeGbif}. Each photograph was also processed to estimate the
percent cover of each type of vegetation using an image processing
software \citep[ImageJ,][]{schneider2012nih}. Hyperspectral reflectance
signatures of each vegetation class were recorded using an ASD FieldSpec
HandHeld 2 spectroradiometer, which acquires reflectance between 325 and
1075 nm, with 1 nm of spectral resolution. Hyperspectral signatures
served dual purposes: they validate the radiometric calibration of drone
data and contribute to misclassification reduction in photo
interpretations.

@@

\caption{\label{fig-vegetation}The five taxonomic classes of vegetation
used to train the Neural Network model and an example of their raw
spectral signatures at the spectral resolution of the Micasense RedEdge
Dual MX. A : Magnoliopsida (\emph{Zostera noltei}) ; B : Phaeophyceae
(\emph{Fucus sp.}) ; C : Rhodophyceae (\emph{Gracilaria
vermiculophylla}) ; D : Chlorophyceae (\emph{Ulva sp.}) ; E :
Bacillariophyceae (Benthic diatoms). \textbf{The taxonomy was chacked
following the World Register of Marine Species (WORMS).}}

@@

A structure-from-motion photogrammetry software \citep[Agisoft
Metashape,][]{agisoft} was used to process images to obtain
multispectral orthomosaics of each flight. The process for
orthomosaicking was identical for every flight. First, key tying points
were detected inside each image and between overlapping images in order
to obtain a sparse point cloud. This cloud was cleaned using a
reprojection accuracy metric to remove noisy points. A dense point cloud
was then produced using a structure from motion algorithm. A surface
interpolation of this dense point cloud was made to obtain a digital
surface model (DSM), used to reconstruct the multispectral ortho-image
\citep{nebel2020review}. Low-altitude drone flights produced
ortho-images with a very high spatial resolution (8 mm per pixel),
making it efficient to visually distinguish between the various types of
vegetation. High-altitude flights allowed to cover larger areas and
produced images with a pixel size of 80 mm (Table~\ref{tbl-flights}).

@@

The spectral similarities of the reflectance signatures at the spectral
resolution of the Micasense senor between intertidal green macrophytes
(Magnoliopsida and Chlorophyceae) make their discrimination challenging
using simple classification algorithms (Figure~\ref{fig-vegetation} F).
To overcome this challenge, a deep learning classification method was
trained, validated, and applied to each drone flight
(Figure~\ref{fig-workflow}).

@@

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./Figures/High_res/Figure3_workflow.png}

@@

\caption{\label{fig-workflow}Schematic representation of the workflow.
Parallelograms represent input or output data, and rectangles represent
Python processing algorithms. The overall workflow of this study is
divided into two distinct parts based on the spatial resolution of the
drone flights: high-resolution flights (pixel size: 8 mm) were used for
training and prediction of the Neural Network model, whereas
lower-resolution flights (pixel size: 80 mm) were solely employed for
prediction purposes. Validation has been performed on both high and
low-resolution flights.}

@@

A dataset containing photo-interpreted drone reflectance pixels was
built to train a Neural Network model. The training pixels were
categorized into seven different classes, representing the various
habitats encountered at the different study sites: sediment, water,
green macroalgae, seagrasses, Benthic diatoms, brown macroalgae and red
macroalgae. Only data from the low-altitude flights
(Table~\ref{tbl-flights}) were used for training because their 8 mm
spatial resolution allowed to avoid spectral sub-pixel mixing and to
accurately identify vegetation classes. \textbf{In the field, seagrasses
displayed two types of color were observed (i.e.~whereas most seagrass
had green leaves, brownish leaves were also observed due to senescence
or photo-degradaration). Careful attention was given to incorporating
training pixels from both color types into the training dataset for the
seagrass class. This approach was consistently applied to all classes
within the model.} More than 418,000 pixels at 8 mm resolution from the
3 training flights were used to train the model
(Table~\ref{tbl-validationPX}). For model training, 21 variables were
used as predictors: the ten raw spectral bands of the Micasense RedEdge
Dual MX multispectral camera (ranging from 444 nm to 840 nm), the same
ten spectral bands standardized using a min/max transformation
(Equation~\ref{eq-std} ; \citep{Cao2017}) and the Normalized difference
vegetation index (NDVI, Equation~\ref{eq-ndvi}). Standardisation of
spectral bands is commonly used to eliminate the scaling differences
between spectra and to limit the effect of biomass on the spectra shape
\citep{Douay2022, Davies2023}.

@@

A neural network classification model was built using the fastai
workflow \citep{howard2018fastai}. This model was composed of 2 hidden
layers and has a total of 26 054 trainable parameters. Parameters have
been fine-tuned using 12 epoch to minimize the error rate. This model
has been called DISCOV, standing for Drone Intertidal Substrat
Classification Of Vegetation.

@@

The workflow of this study revolves around two distinct flight heights
(12 and 120 m, Figure~\ref{fig-workflow}) where ensuring consistency
between reflectances at both heights is crucial. This comparison was
conducted at sites where low and high-altitude flights overlapped. To
compare reflectances of both flights, the low-altitude flights were
resampled to the same spatial resolution and grid as the high-altitude
flights using a median resampling method. Reflectance values were then
extracted, and a scatterplot was generated. The Root Mean Square Error
(RMSE) was computed to compare the difference between the raw and
standardised reflectance.

@@

Variable Importance Plots (VIP) serve as a method to identify which
predictors are important for predicting a specific class. Out of the 21
predictors used in this study, Variable Importance was computed only for
the raw and standardized values of the 10 spectral bands captured by the
MicaSense camera. This is achieved by repeatedly predicting the same
dataset while randomly shuffling one predictor at a time. The benchmark
score obtained after each iteration is then compared to the benchmark
score obtained without shuffling any variables. The greater the
difference between these two benchmark values, the more important the
variable is for the model \citep{WEI2015399}.

@@

We used a Generalized Linear Model (GLM) with a Beta distribution to
examine the relationship between pixel resolution, vegetation class, and
their interaction on the loss of vegetation. The loss of vegetation was
modelled as function of the interaction between pixel resolution and
vegetation class (Benthic diatoms, brown macroalgae, seagrass, green
macroalgae and red macroalgae). Sample vs fitted residuals and
quartile-quartile graphics were assessed visually, to ensure assumptions
of the models used were met.

@@

The key aspect of the workflow adopted in the present study is the
mapping at two different altitudes (12 and 120 m), resulting in two
distinct resolutions for the same area (8 and 80 mm; respectively). The
high-resolution flight was used to estimate the sub-pixel composition
for each pixel of the lower-resolution flight. Consequently, within each
pixel of the high-altitude flights, the contribution of each vegetation
class (\% cover) was obtained, and a kernel density plot was generated.
This plot provided a visual representation of the model's behavior in
mixed vegetation scenarios. It helped to understand the minimum
vegetation cover of a given class within a pixel necessary for the model
to confidently predict that class.

@@

In this study, drone flights were conducted at two different altitudes
(12 and 120 m) to construct the neural network model. At the sites where
the flights at both altitudes overlapped, the reflectance was compared.
Overall there was a good agreement between the two altitudes (RMSE :
0.027 ; Figure~\ref{fig-CompareRef}).

@@

There was a slight underestimation of raw reflectance values in the
high-altitude flight, particularly for higher reflectance values
(Figure~\ref{fig-CompareRef} A). Since both flights were conducted over
vegetated areas, the highest reflectance values correspond to the
infrared part of the spectrum. This difference was not present when the
reflectance has been standardized (Equation~\ref{eq-std} ;
Figure~\ref{fig-CompareRef} B).

@@

Each drone flight was used to produce a prediction map, as well as a
probability map that indicates the model-derived probability of the
selected class for every pixel. The low-altitude flight conducted in
Gafanha, Portugal, represented the site with the highest complexity
(Figure~\ref{fig-GafLow}). Among the five vegetation classes on which
the model was trained, four were present on this site, with green and
red macroalgae mixed with a seagrass meadow. There were also benthic
diatoms biofilms on sediment surface. Although the seagrass was solely
composed of a single species, \emph{Zostera noltei}, two colors of this
species could be observed: dark green (corresponding to healthy leaves)
and brown (when leaves are senescent or have an altered pigment
composition). Regardless of the variation of color, the class
Magnoliopsida (seagrass) was accurately predicted by the model (F1 score
of 0.96 at that site).

@@

\caption{\label{fig-GafLow}RGB orthomosaic (Left) and Prediction (Right)
of the low altitude flight of Gafanha, Portugal. The total extent of
this flight was 3000 m² with a resolution of 8 mm per pixel. The zoom
covers an area equivalent to a 10-meter Sentinel-2 pixel size.}

@@

The high-altitude flight over Gafanha covered a total area of
\textasciitilde1 km² (Figure~\ref{fig-GafHigh}). A channel contouring a
small island was masked in the prediction map. Most of the vegetation
area was classified as seagrass by the model, including patches with
brown leaves. Only a few pixels were classified as green macroalgae (F1
score of 0.55). Patches of red macroalgae were correctly classified (F1
score of 0.85). In the northern part of the site and near the land
edges, patches of the schorre angiosperm \emph{Sporobolus maritimus}
(syn. \emph{Spartina maritima)} were misclassified, either as seagrass
or as brown algae (F1 score of 0.77 and 0.71, respectively).

@@

\caption{\label{fig-GafHigh}RGB orthomosaic (Left) and Prediction
(Right) of the high-altitude flight of Gafanha, Portugal. The total
extent of this flight was about 1 km² with a resolution of 80 mm per
pixel. The yellow outline shows the extent of the low-altitude flight of
Gafanha presented in Figure~\ref{fig-GafLow}. The zoom covers an area
equivalent to a 10-meter Sentinel-2 pixel size.}

@@

Among the high altitude flights, the one acquired over the inner part of
Ria de Aveiro coastal lagoon covered the largest area with approximately
1.5 km² (Figure~\ref{fig-Boat}). The vegetation present at the site was
dominated by seagrass and red macroalgae. The classification provided
consistent results, with a patchy seagrass meadow mixed with red
macroalgae on the eastern part of the site. As shown in the zoom
(Figure~\ref{fig-Boat}), the edges of the meadow were mixed with green
macroalgae (\emph{Ulva sp.}), which the model agreed with (F1 score of
0.89 for green algae, 0.97 for seagrass and 0.98 for red algae).

@@

\caption{\label{fig-Boat}RGB orthomosaic (Top) and Prediction (Bottom)
of the flight made in the inner part of Ria de Aveiro Lagoon, Portugal.
The total extent of this flight was about 1.5 km² with a resolution of
80 mm per pixel. The zoom inserts cover an area equivalent to the size
of a 10-meter Sentinel-2 pixel.}
